Community biomass and species richness temporal trends reveal contrasting assembly and disassembly processes in natural food-webs

### 2022-10-31 paper/manuscript.Rmd
Update results

```{r, results = FALSE}
library(drake)
library(tidyverse)
library(magrittr)
library(cowplot)
library(viridis)
library(kableExtra)
library(sf)
mypath <- rprojroot::find_package_root_file
source(mypath("R/variable_shortcut.R")) # Define your custom code as a bunch of functions.
source(mypath("R/misc.R"))
source_dir(get_mypath("R")) # Define your custom code as a bunch of functions.

@@

Global diversity is reported to decline at alarming rates [@]. At local scale,
recent meta-analysis reported a huge variation in species richness temporal
trends across various taxa and ecosystem types[@vellend_global_2013;
@dornelas_assemblage_2014; @blowes_geography_2019], with no net variations in
species richness[@dornelas_assemblage_2014; @blowes_geography_2019].
Futhermore, recent studies founs that temporal changes in species richness are
unrelated to changes in species composition[@dornelas_assemblage_2014;
@blowes_geography_2019], rising the recognition of the multidimensional nature
of biodiversity changes[@hillebrand_biodiversity_2018; @blowes_local_no_date]
and the need for further investigations on how biodiversity changes over time.

Species interactions among species, described as ecological networks, drive for
a large part community composition, species coexistence, and ecosystem
functioning [@chesson_mechanisms_2000; @thebault_food-web_2003;
@oconnor_general_2017]. The structure of ecological networks can trigger or
impede perturbation propagations, i.e secondary extinctions[@paine_note_1969;
@dunne_network_2002; @srinivasan_response_2007;
@stouffer_comparmentalization_2011]. Higher connectance, which measures the
degree of generalism in a network, is predicted to increase robustness to
secondary extinctions [@dunne_network_2002; @sole_complexity_2001] and studies
found that higher trophic level was associated to a lower temporal variability
of community biomass in theoretical models and empirical
data[@shanafelt_stability_2018; @danet_species_nodate]. Then, accounting for how
the structure of ecological networks changes over time according to other
biodiversity facets is crucial.

Th availability of ecological networks timeseries is
scarce[@kortsch_disentangling_2021], such as predictions about how the strucure
of ecological networks change according to changes in species richness come from
spatial comparison. Spatially, connectance has been shown to be negatively
related to species richness in small (i.e. below 20 species) empirical food-webs
[reviewed in @dunne_network_2006], which reflects constraints on the number of
realizable species interactions or an increasing proportion of specialists in
more rich foodwebs. Connectance has been shown to scale positively or negatively
with community biomass in theoretical food-webs [@thebault_food-web_2003;
@poisot_trophic_2013] while the average trophic level was previously shown to
scale negatively with species richness [@danet_species_2021].

However, predicting how ecological networks will change in time according to
other biodiversity facets based on spatial comparison of ecological communities
is valid as long as assumption of the space-for-time substitution is valid as
well. Space-for-time substitution supposes a homology between variations in
space and in time, i.e. that spatial sequences correspond to temporal sequences.
This method have raised criticisms [@damgaard_critique_2019], with studies
showing analogous patterns in space and time [@blois_space_2013] or not
[@johnson_testing_2008; @adler_contrasting_2007]. Spatial turnover of species
composition, dispersal limitation, disturbance and stochastic processes have
been hypothesised to be responsible for the non-compliance of space-for-time
substitution [@adler_contrasting_2007; @blois_space_2013].

The increasing availability ecological temporal series and the development of
methods to reconstruct food-webs [@bonnaffe_comparison_2021;
@gravel_inferring_2013; @danet_species_2021] makes possible to overcome
those research gaps. We used a standardized monitoring of fish riverine 
communities on the period 1995-2018 over France, documenting the individual size
of fishes in 400 sites during ten years or more. We restructed around 4000
food-webs based on allometric relationships and food diet data, and then
computed connectance, and average trophic level as well as community biomass and
species richness.

@@

```{r}
tar_load(r2_inla_tab)
r2_inla_report <- r2_inla %>%
  select(response, r2_mvp_marg95hpdci, r2_mvp_cond95hpdci)
r2ic <- map(
  setNames(r2_inla_report$response, r2_inla_report$response),
  ~r2_inla_report[r2_inla_report$response == .x, ]$r2_mvp_cond95hpdci
)

tar_load(r2_sem)
```
Using a hierarchical model to estimate conjointly the temporal trends of  
foodweb strucure at each site, we found weak evidence of a global decrease in
community biomass (CI80%:`r gbm` by decade; median [Credible Interval]) but of a
global increase in species richness (CI80%: `r gr` by decade) across the `r nb_st` sites. 
We did not find evidence of a global trends in connectance and average trophic
level (CI80%: `r gct` and `r gt` respectively). The hierarchical model captured
well the variance associated to the spatial structure of the data with
conditional $R^2$ going from  `r r2ic$log_bm_std` (mean [CI95%]) for community
biomass to `r r2ic$log_rich_std` for species richness (Table SXX).

@@

We found a strong positive relationship between the temporal trends of community
biomass and the ones of species richness ($r_\delta = `r rdtps$bm$r`$,
Fig. \@ref(fig:sem) b), meaning that both temporal trends
had the same direction in average. At the sites 
where species richness decreased over time, community biomass also decreased
over time. On the other hand, at the sites where species richness increased over
time, community biomass also increased over time. Along the spatial gradient, we also
found a positive relationship between median community biomass and median
species richness, but half the strength as the temporal one
($r_\delta = `r rdsp$bm$r`$).

The temporal trends of connectance were positively linked with the ones of
community biomass ($r_\delta = `r rdtps$ct$bm`$), also meaning that both
temporal trends had the same direction in average. That is, the sites where
food-web connectance decreased over time also had community biomass that
decreased over time and reversely. It suggests that communities that decreased
in biomass over time loose generalist predators over time, as connectance
measures the degree of generalism, and reversely that communities
gaining biomass over time also gain generalist predators. Spatially, we also found
a positive relationship between median connectance and median species
richness ($r_\delta = `r rdsp$ct$bm`$). We found a negative 
relationship between the temporal trends of connectance and the
ones of species richness ($r_\delta = `r rdtps$ct$r`$, Fig. \@ref(fig:sem)d)
while we found negative relationship three times stronger between median connectance and
median species richness along the spatial gradient 
($r_\delta = `r rdsp$ct$r`$, Fig. \@ref(fig:std)).

The temporal trends of average trophic level had a strong positive
relationship with the ones of community biomass 
($r_\delta = `r rdtps$tlvl$bm`$, Fig. \@ref(fig:sem)B), meaning
that sites where community biomass decreased over time, average trophic level
decreased over time and *vice versa*. It means that in communities loosing biomass
over time, the average biomass of the food-web moved down to lower trophic level
and *vice versa*. We also found a positive relationship between
median average trophic level and median community biomass in space, but half the
strength as the temporal one ($r_\delta = `r rdsp$tlvl$bm`$, Fig.
\@ref(fig:std)). The temporal trends of average trophic level was unrelated to
the ones of species richness. It suggests that in communities either loosing or
gaining species richness over time, the location of average biomass along
trophic level did not significantly change according to species richnnes.
However, we found a strong negative relationship in space 
($r_\delta = `r rdsp$tlvl$r`$, Fig. \@ref(fig:std)).

@@

The structural model on the temporal trends of food-web structure reveals that
the species richness has a direct negative effect on connectance but also a
positive one through community biomass
($r_\delta$ = `r ridtps$ct$r`). It resulted a relatively weak total negative
effect of species richness on connectance (`r rttps$ct$r`). Spatially,
we also found a positive indirect effect of richness on connectance through
community biomass but a weak one (`r ridsp$ct$r`), resulting in a strong negative total
effect of richness on connectance (`r rtsp$ct$r`).

The relationships among temporal trends showed a indirect positive effect of
species richness on average trophic level (`r ridtps$tlvl$r`), resulting in a total positive
effect of species richness on average trophic level (`r rttps$tlvl$r`). Over the spatial
gradient, we found the same positive indirect effect of richness through
community biomass (`r ridsp$tlvl$r`). WHATTTT? But as the direct effect of
richness on average trophic level was strongly negative, it resulted in a total
strong negative effect of species richness on average trophic level over the
spatial gradient (`r rtsp$tlvl$r`).

@@

@@

Most temporal trends of for species richness and community biomasss were
classified as stable. Roughly the same number of stations displayed temporal
decrease and increase for species richness, and twice more stations displayed
decrease in community biomass than increase. This result is in accordance with
previous meta analysis on species richness trends of vertebrate in marine,
freshwater and terrestrial systems, which show no change in species richness in
average [@dornelas_assemblage_2014; @blowes_geography_2019] but an increase in
total abundance [@klink_meta-analysis_2020].

@@

The temporal trends of species richness had an indirect positive effect on the
temporal trends of both connectance and average trophic level through the
positive relationship between the temporal trends of species richness and
community biomass. The temporal trends of species richness had then an
total effect on the temporal trends of connectance and average trophic level
which was twice lower in magnitude than the magnitude of the total effect of the
temporal trends of community biomass. It suggests that the temporal assembly and
disassembly of food-webs are more sensible to variations in community biomass
than in species richness, in other words that food-web structure is more robust
to species richness change than biomass changes. This result is in line with
evidences that food-web containing generalist predators can rewire in response
to changes in the availability of prey species [@bartley_food_2019]. *(J'ai
regardé dans la biblio envoyé par Elisa et j'ai refais de la biblio moi-même, je
n'ai rien qui s'approche de ce résultat).* This result also resonates with the
energy paradigm of food-webs which proposes that food-webs are primarily
constrained by fluxes of matter and energy [@lindeman_trophic-dynamic_1942], so
community biomass and not by species richness.

- Remove the following paragraph:

As studies reported community biomass or species richness temporal trends
[@hallmann_more_2017; @ceballos_biological_2017; @dornelas_assemblage_2014;
@blowes_geography_2019], we document that temporal trends of species richness
and community biomass might have different consequences on the structure of the
community. In particular, the decrease of the average trophic level associated
to the loss of community biomass could have cascading effect on ecosystem
functioning as the highest predators are predicted to increase productivity and
stabilise communities [@wang_biodiversity_2018; @shanafelt_stability_2018;
@danet_species_nodate]. In contrasts, the loss of species richness might be
linked to loss of redundancy in the food-web, highlighting that those
communities may become more at risk regarding future perturbations
[@yachi_biodiversity_1999]. Interestingly, the hypothesis that temporal trends
in community biomass is related to higher trophic level in our system has a
strong reasoning with the effect of human fisheries in marine systems which
affect primarily fish species at top trophic level [@sprules_surfing_2016].
Overall, our study highlights that monitoring several facets of biodiversity is
essential to infer how ecological communities are reshaped by loss and gain of
biomass and species richness.

@@

Surprisingly, we found that the temporal trends of connectance were weakly
related to the one of species richness as found spatially in most empirical
food-webs [@dunne_network_2006] and also here. We further found a weak negative
relationship between the temporal trends of average trophic level and the one of
species richness which we found a twice stronger but positive relationship
between median average trophic level and median species richness over the
spatial gradient. An explanation might be that environmental gradient, in
resources in hydrologic contraints might be responsible for the strong negative
effect of species richness on both connectance and average trophic level over
spatial gradient. However, we controlled our spatial analysis by the Biological
Demand in Oxygen, water temperature and the stream gradient. This result is also
in line with other studies which found mismatch between spatial and temporal
patterns [@adler_contrasting_2007; @johnson_testing_2008]. This discrepancy in
space and time might be due to the crucial assumption that variations in space
reflect differences in assembly stages of ecological communities and then
mirrors their temporal assembly/disassembly trajectories. Temporal variation of
species richness might relate on transitory dynamics while spatial gradient
reflect aggregated dynamics over longer time scale including a far number of
processes including high species turnover along spatial gradient. Another reason
might be that the spatial gradient displays a higher range of variation for
species richness than the temporal stations (Fig. S2). However, we found
temporal trends of species richness of -70%/-50% to +80%/+100% by decade.

- Développer un point la-dessus?:

Our study highlights that time series are an important tool to
understand the assembly and disassembly of food-webs following changes in
species richness and biomass. Our study also highlights the usefulness of
structural equation model to highlight direct and indirect pathways.

---
### Biomass and richness gains and losses are symmetric processes in terms of community structure
#
#We found no case in which gain or losses of biomass and richness lead to
#different relationship between the temporal dynamic of community structure and
#the one of biomass and species richness. We could have imagined that
#extinction/colonisation processes or immigration/extinction debt could have lead
to different relationships.
---

>>>>

```{r, results = FALSE}
library(targets)
library(tidyverse)
library(magrittr)
library(cowplot)
library(viridis)
library(kableExtra)
library(sf)
mypath <- rprojroot::find_package_root_file
source(mypath("R/variable_shortcut.R")) # Define your custom code as a bunch of functions.
source(mypath("R/misc.R"))
source_dir(get_mypath("R")) # Define your custom code as a bunch of functions.

@@

Global diversity is reported to decline at alarming rates [@]. At local scale,
recent meta-analysis reported a huge variation in species richness temporal
trends across various taxa and ecosystem types[@vellend_global_2013;
@dornelas_assemblage_2014; @blowes_geography_2019], with no net variations in
species richness[@dornelas_assemblage_2014; @blowes_geography_2019].
Futhermore, recent studies founs that temporal changes in species richness are
unrelated to changes in species composition[@dornelas_assemblage_2014;
@blowes_geography_2019], rising the recognition of the multidimensional nature
of biodiversity changes[@hillebrand_biodiversity_2018; @blowes_local_nodate]
and the need for further investigations on how biodiversity changes over time.

Species interactions among species, described as ecological networks, drive for
a large part community composition, species coexistence, and ecosystem
functioning[@chesson_mechanisms_2000; @thebault_food-web_2003;
@oconnor_general_2017]. The structure of ecological networks can trigger or
impede perturbation propagations, i.e secondary extinctions[@paine_note_1969;
@dunne_network_2002; @srinivasan_response_2007;
@stouffer_compartmentalization_2011]. Higher connectance, which measures the
degree of generalism in a network, is predicted to increase robustness to
secondary extinctions[@dunne_network_2002; @sole_complexity_2001] and studies
found that higher trophic level was associated to a lower temporal variability
of community biomass in theoretical models and empirical
data[@shanafelt_stability_2018; @danet_species_2021]. Then, accounting for how
the structure of ecological networks changes over time according to other
biodiversity facets is crucial.

Th availability of ecological networks timeseries is
scarce[@kortsch_disentangling_2021], such as predictions about how the strucure
of ecological networks change according to changes in species richness come from
spatial comparison. Spatially, connectance has been shown to be negatively
related to species richness in small (i.e. below 20 species) empirical food-webs
[reviewed in @dunne_network_2006], which reflects constraints on the number of
realizable species interactions or an increasing proportion of specialists in
more rich foodwebs. Connectance has been shown to scale positively or negatively
with community biomass in theoretical food-webs[@thebault_food-web_2003;
@poisot_trophic_2013] while the average trophic level was previously shown to
scale negatively with species richness[@danet_species_2021].

However, predicting how ecological networks will change in time according to
other biodiversity facets based on spatial comparison of ecological communities
is valid as long as assumption of the space-for-time substitution is valid as
well. Space-for-time substitution supposes a homology between variations in
space and in time, i.e. that spatial sequences correspond to temporal sequences.
This method have raised criticisms[@damgaard_critique_2019], with studies
showing analogous patterns in space and time[@blois_space_2013] or not
[@johnson_testing_2008; @adler_contrasting_2007]. Spatial turnover of species
composition, dispersal limitation, disturbance and stochastic processes have
been hypothesised to be responsible for the non-compliance of space-for-time
substitution[@adler_contrasting_2007; @blois_space_2013].

The increasing availability ecological temporal series and the development of
methods to reconstruct food-webs[@bonnaffe_comparison_2021;
@gravel_inferring_2013; @danet_species_2021] makes possible to overcome
those research gaps. We used a standardized monitoring of fish riverine 
communities on the period 1995-2018 over France, documenting the individual size
of fishes in 400 sites during ten years or more. We restructed around 4000
food-webs based on allometric relationships and food diet data, and then
computed connectance, and average trophic level as well as community biomass and
species richness.

@@

```{r}
tar_load(c(r2_inla_tab, r2_inla, r2_sem))
r2_inla_report <- r2_inla %>%
  select(response, r2_mvp_marg95hpdci, r2_mvp_cond95hpdci)
r2ic <- map(
  setNames(r2_inla_report$response, r2_inla_report$response),
  ~r2_inla_report[r2_inla_report$response == .x, ]$r2_mvp_cond95hpdci
)
```

Using a hierarchical model to estimate conjointly the temporal trends of  
foodweb strucure at each site, we found weak evidence of a global decrease in
community biomass (CI80%:`r gbm` by decade; median [Credible Interval]) but of a
global increase in species richness (CI80%: `r gr` by decade) across the `r nb_st` sites. 
We did not find evidence of a global trends in connectance and average trophic
level (CI80%: `r gct` and `r gt` respectively). The hierarchical model captured
well the variance associated to the spatial structure of the data with
conditional $R^2$ going from  `r r2ic$log_bm_std` (mean [CI95%]) for community
biomass to `r r2ic$log_rich_std` for species richness (Table SXX).

@@

@@

Using structural equation models to disentangle the direct and indirect
relationships among foodweb structure descriptors, we found a strong positive
relationship between the temporal trends of community biomass and of
species richness ($r_\delta = `r rdtps$bm$r`$, Fig. \@ref(fig:sem) a and b,
$r_\delta$: standardized slope coefficients, [95% CI]), meaning that both
temporal trends had the same direction in average. At the sites where species
richness decreased over time, community biomass also decreased over time. On the
other hand, at the sites where species richness increased over time, community
biomass also increased over time.

The temporal trends of connectance were positively linked with the ones of
community biomass ($r_\delta = `r rdtps$ct$bm`$), also meaning that both
temporal trends had the same direction in average. That is, the sites where
food-web connectance decreased over time also had community biomass that
decreased over time and reversely. It suggests that communities that decreased
in biomass over time lose highly connected trophic species over time, and
reversely that communities gaining biomass over time also gain highly connected
trophic species over time. We found a negative relationship between the temporal
trends of connectance and the ones of species richness
($r_\delta = `r rdtps$ct$r`$, Fig. \@ref(fig:sem)d). We further found the
species richness had an indirect positive effect on connectance, through
community biomass ($r_\delta = `r ridtps$ct$r`$). It resulted in a relatively
weak total negative effect of species richness on connectance
($r_\delta = `r rttps$ct$r`$).

The temporal trends of average trophic level had a strong positive relationship
with the ones of community biomass ($r_\delta = `r rdtps$tlvl$bm`$, Fig.
\@ref(fig:sem)e), meaning that sites where
community biomass decreased over time, average trophic level decreased over time
and reversely. It means that in communities losing biomass over time, the
average biomass of the food-web moved down to lower trophic level and vice
versa. However, we found that the temporal trends of average trophic level had
no direct relationship the ones of species richness ($r_\delta = `r rdtps$tlvl$r`$,
Fig. \@ref(fig:sem)f). It suggests that in communities either
losing or gaining species richness over time, the location of average biomass
along trophic level does not significantly change according to species richness.
However, the temporal trends of species richness had an indirect positive
relationship with the temporal trends of average trophic level through community
biomass ($r_\delta = `r ridtps$tlvl$r`$), resulting in a total positive
relationship ($r_\delta = `r rttps$tlvl$r`$).

@@

As for the temporal analysis, we found a positive relationship between
median community biomass and median species richness along the spatial gradient,
but half the strength as the temporal one ($r_\delta = `r rdsp$bm$r`$, Fig.
\@ref(fig:std)). We further found a positive relationship between median
connectance and median community biomass ($r_\delta = `r rdsp$ct$bm`$), as in
temporal trends analysis. However, we found a negative relationship between
median connectance and median species richness, but
three times stronger than in the temporal analysis ($r_\delta = `r rdsp$ct$r`$,
Fig. \@ref(fig:std)). We also found a positive indirect effect of median
richness on median connectance through median community biomass but a weak one
($r_\delta = `r ridsp$ct$r`$), resulting in a strong negative total effect of
median richness on median connectance ($r_\delta = `r rtsp$ct$r`$).

We found a positive relationship between median average trophic level and median
community biomass, but half the strength as in the temporal analysis ($r_\delta
= `r rdsp$tlvl$bm`$, Fig. \@ref(fig:std)). However, we found a strong negative
relationship between median average trophic level and median richness ($r_\delta
= `r rdsp$tlvl$r`$), in contrast with findings in temporal analysis. We also
found a positive indirect relationship between average trophic level and  median
richness through median community biomass (`r ridsp$tlvl$r`), resulting in a
strong total negative effect of median richness on median average trophic level
(`r rtsp$tlvl$r`).

@@

Most temporal trends of for species richness and community biomasss were
classified as stable. Roughly the same number of stations displayed temporal
decrease and increase for species richness, and twice more stations displayed
decrease in community biomass than increase. This result is in accordance with
previous meta analysis on species richness trends of vertebrate in marine,
freshwater and terrestrial systems, which show no change in species richness in
average [@dornelas_assemblage_2014; @blowes_geography_2019] but an increase in
total abundance [@van_klink_meta-analysis_2020].

@@

The temporal trends of species richness had an indirect positive effect on the
temporal trends of both connectance and average trophic level through the
positive relationship between the temporal trends of species richness and
community biomass. The temporal trends of species richness had then an
total effect on the temporal trends of connectance and average trophic level
which was twice lower in magnitude than the magnitude of the total effect of the
temporal trends of community biomass. It suggests that the temporal assembly and
disassembly of food-webs are more sensible to variations in community biomass
than in species richness, in other words that food-web structure is more robust
to species richness change than biomass changes. This result is in line with
evidences that food-web containing generalist predators can rewire in response
to changes in the availability of prey species [@bartley_food_2019]. This result
also resonates with the energy paradigm of food-webs which proposes that
food-webs are primarily constrained by fluxes of matter and energy
[@lindeman_trophic-dynamic_1942], so community biomass and not by species
richness.

---
# - Remove the following paragraph:  
# As studies reported community biomass or species richness temporal trends
# [@hallmann_more_2017; @ceballos_biological_2017; @dornelas_assemblage_2014;
# @blowes_geography_2019], we document that temporal trends of species richness
# and community biomass might have different consequences on the structure of the
# community. In particular, the decrease of the average trophic level associated
# to the loss of community biomass could have cascading effect on ecosystem
# functioning as the highest predators are predicted to increase productivity and
# stabilise communities [@wang_biodiversity_2018; @shanafelt_stability_2018;
# @danet_species_nodate]. In contrasts, the loss of species richness might be
# linked to loss of redundancy in the food-web, highlighting that those
# communities may become more at risk regarding future perturbations
# [@yachi_biodiversity_1999]. Interestingly, the hypothesis that temporal trends
# in community biomass is related to higher trophic level in our system has a
# strong reasoning with the effect of human fisheries in marine systems which
# affect primarily fish species at top trophic level [@sprules_surfing_2016].
# Overall, our study highlights that monitoring several facets of biodiversity is
# essential to infer how ecological communities are reshaped by loss and gain of
# biomass and species richness.
---

@@

Surprisingly, we found that the temporal trends of connectance were weakly
related to the one of species richness as found spatially in most empirical
food-webs [@dunne_network_2006] and also here. We further found a weak negative
relationship between the temporal trends of average trophic level and the one of
species richness which we found a twice stronger but positive relationship
between median average trophic level and median species richness over the
spatial gradient. An explanation might be that environmental gradient, in
resources in hydrologic contraints might be responsible for the strong negative
effect of species richness on both connectance and average trophic level over
spatial gradient. However, we controlled our spatial analysis by the Biological
Demand in Oxygen, water temperature and the stream gradient. This result is also
in line with other studies which found mismatch between spatial and temporal
patterns [@adler_contrasting_2007; @johnson_testing_2008]. This discrepancy in
space and time might be due to the crucial assumption that variations in space
reflect differences in assembly stages of ecological communities and then
mirrors their temporal assembly/disassembly trajectories. Temporal variation of
species richness might relate on transitory dynamics while spatial gradient
reflect aggregated dynamics over longer time scale including a far number of
processes including high species turnover along spatial gradient.
